Regionally Additive Models: Explainable-by-design models minimizing feature interactions

### 2023-06-22 report/report.tex
intro updated

% Paragraph for Motivating about Regionally Additive Models
Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models.
Their popularity stems from their seamless interpretability; since they are a linear (additive) combination of univariate functions,
\(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\), each individual univariate function (component) can be readily visualized and comprehended in isolation.
However, GAM's main limitation is that they cannot express interactions between features.
To mitigation this limitation, some approaches~\citep{lou2013accurate} extend them enabling pairwise interactions,
i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{c \neq s} f_{sc}(x_s, x_c)\).
Pairwise interactions can also be visualized and understood in isolation, so these models also maintain their x-by-design nature.
Unfortunately, this does not hold for any interaction that involves more than two features, thus, the expressiveness of GAMs is limited to capturing up to two-feature interactions.

% Paragraph for Motivating about Regionally Additive Models
To overcome this limitation, we propose Regionally Additive Models (RAMs), a novel class of x-by-design models,
that fits multiple GAMs to subregions of the feature space where interactions are minimized.
Our approach consists of a three-step pipeline.
First, we fit a black-box model to capture all high-order interactions.
Second, we identify the subregions where the black-box model is nearly locally additive.
Finally, we train a GAM specifically for each identified subregion.

>>>>

% Paragraph for Motivating about Regionally Additive Models
Generalized Additive Models (GAMs)~\citep{hastie1987generalized} are a popular class of explainable by design (x-by-design) models~\citep{rudin2019stop, ghassemi2021false}.
Their popularity stems from their inherent interpretability.
GAMs represent an aggregation of univariate functions, where the overall model can be expressed as \(f(\xb) = c + \sum_{s=1}^D f_s(x_s)\).
Due to this structure, each individual univariate function (component) can be visualized and interpreted independently.
Consequently, understanding the behavior of the overall model simply requires visualizing all components, each with a one-dimensional plot.

However, GAMs encounter limitations when predicting outcomes that depend on multiple features simultaneously.
In such cases, the unknown predictive function includes terms representing interactions between features.
To mitigate this limitation, GA$^2$Ms~\citep{lou2013accurate} extended GAMs adding pairwise interactions in the
formulation, i.e., \(f(\xb) = c + \sum_{s=1}^D f_s(x_s) + \sum_{s=1}^D \sum_{s_1 \neq s_2} f_{s_1 s_2}(x_{s_1}, x_{s_2})\).
Interpreting GA$^2$Ms involves visualizing both the first-order ($1D$ plots) and second-order (($2D$ plots)) components.
However, as the number of features increases, the number of pairwise interactions grows exponentially,
making it impractical for users to remember and interpret a large number of two-dimensional plots.
Therefore, these methods target on automatically selecting the most significant interaction terms.

Both GAMs and GA$^2$Ms have limitations in modeling interactions beyond the second degree,
and as far as we know, there is no existing work that focuses on x-by-design GAM-based models capable of
representing higher-order interactions.
The main challenge lies in the difficulty of visualizing more than three terms simultaneously,
which would violate the x-by-design principle.
To address this limitation, we propose a new class of x-by-design models called Regionally Additive Models (RAMs).

RAMs tackle the challenge of visualizing general interaction terms involving more than two variables by focusing
on a specific class of terms.
These terms have a general structure of these terms is
$f(x_{s_1} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for first-degree interactions and
$f(x_{s_1}, x_{s_2} | \when{x_{c_1}}, \when{x_{c_2}}, \cdots)$ for second-degree interactions.
To better grasp the idea, consider a prediction task where the outcome depends, among others,
on a combination of three features: $x_1 \in [20, 80]$ (age), $x_2 \in [0, 40]$ (years in work), and $x_3 \in \{True, False\}$ (married).
Both GAM and GA$^2$M would fail to capture the effect of the three-feature interaction.
However, the three-feature effect can be decomposed in two sets of second-degree conditional terms based on the marital status:
$f_{1}(x_1, x_2 | x_3 = True)$
and
$f_{2}(x_1, x_2 | x_3 = False)$, which can be visualized and interpreted as using two-dimensional plots.
In a different scenario, it may be better to it in four sets of first-degree conditional terms:
$f_{1}(x_1 | x_2 < 10, x_3 = True)$,
$f_{2}(x_1 | x_2 \geq 10, x_3 = True)$,
$f_{3}(x_1 | x_2 < 10, x_3 = False)$, and
$f_{4}(x_1 | x_2 \geq 10, x_3 = False)$, which can be visualized and interpreted as four one-dimensional plots.
% Paragraph for Motivating about Regionally Additive Models

The objective of RAMs is to automatically identify the most significant conditional terms and fit a GAM for each of them.
The RAM framework involves three steps.
First, a black-box model is fitted to capture all high-order interactions.
Then, the subregions where the black-box model exhibits near-local additivity are identified using Regional Effect Plots (REPs).
Finally, a GAM is trained specifically for each identified subregion.
